---
title: "Healthy Donor Seurat"
author: "Emma Nguyen"
date: "2024-06-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = normalizePath("C:/Emma /healthy donors"))
```

# Install necessary packages
```{r}
#install.packages("Seurat")
#packageVersion("Seurat")

library(SingleCellExperiment)
library(Seurat)
library(tidyverse)
library(Matrix)
library(scales)
library(cowplot)
library(RCurl)
library(dplyr)
library(ggplot2)
library(patchwork)
library(Matrix)
```

```{r}
work_dir <- ("~/Emma /healthy donors_GSE224995")
setwd(work_dir)
list.files()
```

# Inspect data
```{r}
library(data.table)
features <- fread("features.tsv.gz")
head(features, 50)

barcodes <- fread("barcodes.tsv.gz")
head(barcodes, 50)

con <-gzfile("matrix.mtx.gz", "rt")
lines <- readLines(con, n=50)
close(con)
cat(lines, sep = "\n")
```

# Load data into a matrix and create Seurat Obj
Attempt 1
```{r}
# set up working path to the files

# if working directory is set up properly, no need for path
#matrix_path <- "/mnt/workspace/Emma/healthy donors_GSE224995/matrix.mtx.gz"
#features_path <- "/mnt/workspace/Emma/healthy donors_GSE224995/features.tsv.gz"
#barcodes_path <- "/mnt/workspace/Emma/healthy donors_GSE224995/barcodes.tsv.gz"

expression_mtx <- ReadMtx(mtx = 'matrix.mtx.gz',features = 'features.tsv.gz', cells = 'barcodes.tsv.gz', feature.column = 1)


obj <- CreateSeuratObject(counts = expression_mtx, project = "heart", 
                          min.cells = 3, min.features = 200)
```

Attempt 2
```{r}
duplicated_rows <- expression_mtx[duplicated(rownames(expression_mtx)),]

expression_mtx <- expression_mtx[!duplicated(rownames(expression_mtx)),]

obj <- CreateSeuratObject(counts = expression_mtx, project = "heart", 
                          min.cells = 3, min.features = 200)


```

Attempt 3
```{r}
matx <- readMM('matrix.mtx.gz')
matx <- Matrix::t(matx)
matx <- as(matx, 'dgCMatrix')

features <- readLines("features.tsv.gz")
barcodes <- readLines("barcodes.tsv.gz")

colnames(matx) <- barcodes
rownames(matx) <- features

assay <-CreateAssayObject(counts= matx, assay = 'RNA', min.features = 200, min.cells =3)
obj <- CreateSeuratObject(assay)
```

Attempt 4 # This works #
```{r}
data_dir <- "~/Emma /healthy donors_GSE224995"
data <- Read10X(data.dir= data_dir, gene.column =1)

seurat_obj <- CreateSeuratObject(counts= data)

# Add metadata to Seurat objects
metadata <- read.table("features.tsv.gz", header = TRUE, sep= "\t")
head(metadata)

seurat_barcodes <- colnames(seurat_obj)
head(seurat_barcodes)

rownames(metadata) <- metadata[,1]
metadata <- metadata[,-1]

# add rownames and colnames to metadata
metadata <- metadata[match(seurat_barcodes, rownames(metadata)),]

head(seurat_obj@meta.data)
```

# Quality Control
# don't need this 
```{r}
# https://github.com/hbctraining/In-depth-NGS-Data-Analysis-Course/blob/master/sessionIV/lessons/SC_quality_control_analysis.md

# Load the dataset
counts <- readMM("matrix.mtx.gz")
genes <- read_tsv("features.tsv.gz", col_names = FALSE)
gene_ids <- genes$X1 
cell_ids <- read_tsv("barcodes.tsv.gz", col_names = FALSE)$X1

# Make column names as the cell IDs and row names as gene IDs
rownames(counts) <- gene_ids
colnames(counts) <- cell_ids

# Create metadata table
metadata <- data.frame(row.names = cell_ids, cells = cell_ids, stringsAsFactors= F)

# Add number of UMIs to meta
metadata$nUMI <- Matrix::colSums(counts)

# Add number of genes detected
metadata$nGene <-  Matrix::colSums(counts >0)

# Add number of UMIs/ gene
metadata$log10GenesPerUMI <-  log10(metadata$nGene)/ log10(metadata$nUMI)
```


# Dowloaded AnnotationHub and Ensembl if needed 
```{r}
#install.packages("BioManager")
#BiocManager::install("AnnotationHub")
#library(AnnotationHub)

#ah <- AnnotationHub()

# Access Ensembl database
#ahDb <- query(ah, pattern = c("Homo sapiens", "EnsDb"), ignore.case = TRUE)

# Extract gene-level information
annotaions <- AnnotationDbi::select(
  org.Hs.eg.db, keys= gene_results$gene_id,
  columns = c('SYMBOL', 'GENENAME'), keytype = 'ENTREZID'
)


# Extract annotations of interest
library(ensembldb)
library(EnsDb.Hsapiens.v86)
edb <- EnsDb.Hsapiens.v86
annotations <- genes(edb, return.type= 'data.frame')
View(annotaions)
  
annotations <- annotations %>%
  dplyr::select(gene_id, gene_name, gene_biotype, seq_name, description, entrezid)

# See biotypes
annotations$gene_biotype %>%
  factor() %>%
  levels()
```


# Visualize QC metrics
```{r}
# do not concern Mitochondrial DNA
VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2)
```

```{r}
# Relationships between count and feature 
plot1 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 
```

# Normalizing the data
```{r}
seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", scale.factor = 10000)
```

# Subset for high cell-cell variation
```{r}
seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", nfeatures = 2000)

# Identify 10 most highly variable genes
top10 <- head(VariableFeatures(seurat_obj), 10)
print(top10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(seurat_obj)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2

plot2
```

# Scaling the data
```{r}
all.genes <- rownames(seurat_obj)
seurat_obj <- ScaleData(seurat_obj, features = all.genes)
```

# PCA
```{r}
seurat_obj <- RunPCA(seurat_obj, features = VariableFeatures(object = seurat_obj))

print(seurat_obj[["pca"]], dims = 1:5, nfeatures = 5)
```

```{r}
VizDimLoadings(seurat_obj, dims = 1:2, reduction = "pca")
```

```{r}
DimPlot(seurat_obj, reduction = "pca") 
```

```{r}
DimHeatmap(seurat_obj, dims = 1, cells = 500, balanced = TRUE)
```

# Determine dimension
```{r}
ElbowPlot(seurat_obj)
```
```{r}
# another way of checking dimension of dataset
dims_av <- length(seurat_obj[["pca"]][[1]])
seurat_ob <- JackStraw(seurat_obj, num.replicate =100)
seurat_ob <- ScoreJackStraw(seurat_ob, dims = 1:20)
JackStrawPlot(seurat_ob, dims = 1:20)
```

# Cluster the cells 
```{r}
seurat_obj <- FindNeighbors(seurat_obj, dims= 1:15) # less variation after 15 as above
seurat_obj <- FindClusters(seurat_obj, resolution =0.5) # 0.4-1.2, increase = more cluster

# Look at cluster IDs of the first 5 cells
head(Idents(seurat_obj), 5)
```


# UMAP/tSNE
```{r}
seurat_obj <- RunUMAP(seurat_obj, dims = 1:15) # 15 as determine above

DimPlot(seurat_obj, reduction = "umap")



saveRDS(seurat_obj, file = "../umap.rds")
```

# Finding cluster biomarkers
```{r}
# find top 5 markers of cluster 2
cluster2.markers <- FindMarkers(seurat_obj, ident.1 = 2)
head(cluster2.markers, n = 5)
```


```{r}
# find markers for only positive cluster, only positive

seurat.markers <- FindAllMarkers(seurat_obj, only.pos = TRUE)
seurat.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)
print(seurat.markers)

# this will list all marker in each cluster, might be a bit hard to see 
markers_table <- seurat.markers %>%
  group_by(cluster) %>%
  summarise(markers = paste(gene, collapse = ",")) %>%
  as.data.frame()
print(markers_table)

# this save all marker into a table, might need to reopen in Excel to get the format right 
write.table(seurat.markers, file= "markers_table.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

```

# Some plot
```{r}
# plot features
VlnPlot(seurat_obj, features = c("RBM47", "CD163"))


# plot raw counts 
VlnPlot(seurat_obj, features = c("RBM47", "CD163"), slot = "counts", log = TRUE)
```

```{r}
# Pick gene names as preferred
# Split into smaller chunks for easier to read graph
FeaturePlot(seurat_obj, features = c("F13A1", "MRC1", "CD163", "SLCO2B1","RBM47"))
                                    
                                    #  "EGFLAM", "GUCY1A2", "NR2F2-AS1", "ENPEP", "ABCC9",
                                    #  "SLC9A9", "GAS7", "EPB41L3", "COLEC12",
                                    # "DOCK2", "BICC1", "FBN1", "DCN", "GLIS3", "NEGR1" ))
```
```{r}
FeaturePlot(seurat_obj, features = c("EGFLAM", "GUCY1A2", "NR2F2-AS1", "ENPEP", "ABCC9"))
```

```{r}
FeaturePlot(seurat_obj, features = c("SLC9A9", "GAS7", "EPB41L3", "COLEC12"))
```

```{r}
FeaturePlot(seurat_obj, features = c("DOCK2", "BICC1", "FBN1", "DCN", "GLIS3", "NEGR1" ))
```

```{r}
# top 2 marker genes for each cluster 
features <- seurat.markers %>%
    group_by(cluster) %>%
  top_n(n=2, wt= avg_log2FC)
print(features)  

write.table(features, file= "top_features.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

---------
# Assigning cell type identity to clusters
# Need cell types reference for each cluster
```{r}
new.cluster.ids <- c("# go to Ethan to find reference")
names(new.cluster.ids) <- levels(seurat_obj)
pbmc <- RenameIdents(seurat_obj, new.cluster.ids)

DimPlot(seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```


```{r}
plot <- DimPlot(seurat_obj, reduction = "umap", label = TRUE, label.size = 4.5) + xlab("UMAP 1") + ylab("UMAP 2") +
    theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + guides(colour = guide_legend(override.aes = list(size = 10)))

ggsave(filename = "../output/images/pbmc3k_umap.jpg", height = 7, width = 12, plot = plot, quality = 50)
```

------
# scCATCH
```{r}
install.packages("scCATCH")
library(scCATCH)

counts <- GetAssayData(seurat_obj, slot = "counts")
seurat_clusters <- seurat_obj$seurat_clusters
head(seurat_clusters)
table(seurat_clusters)

dim(counts)
dim(seurat_clusters)

?createscCATCH
seurat_clusters <- as.character(seurat_clusters)
```


```{r}
# create a scCATCH object
sc_obj <- createscCATCH(data= counts, cluster = seurat_clusters)
str(sc_obj)

# need to check "cellmatch" database
data(cellmatch)
str(cellmatch)
unique(cellmatch$species)
unique(cellmatch$tissue)

human_heart_marker <- cellmatch[cellmatch$species== "Human" & cellmatch$tissue== "Heart",]
view(human_heart_marker)

# find marker gene for each cluster
sc_obj <- findmarkergene(object= sc_obj, species = "Human", marker= cellmatch, tissue= "Heart")
head(sc_obj@marker)

sc_obj <- findcelltype(sc_obj)
sc_obj@celltype

write.table(sc_obj@celltype, file= "cell_type.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```


------
# Volcano Plot 
```{r}
install.packages("harmony")
library(harmony)
library(RColorBrewer)
library(ggrepel)
library(gridExtra)

BiocManager::install('EnhancedVolcano')
BiocManager::install('DESeq2')
library(EnhancedVolcano)
library(readr)
library(DESeq2)
library(AnnotationDbi)
library("org.Hs.eg.db")
library(ComplexHeatmap)
library(circlize)

BiocManager::install('clusterProfiler')
library(gprofiler2)
library(clusterProfiler)
```

```{r}
min(seurat.markers$avg_log2FC)
max(seurat.markers$avg_log2FC)
min(seurat.markers$p_val_adj)
max(seurat.markers$p_val_adj)

#  adjusted p = 0.05, FC= 2.5,
EnhancedVolcano(
  seurat.markers,
  lab= rownames(seurat.markers),
  x= "avg_log2FC",
  y= "p_val_adj",
  title = "Enhanced Volcano Plot",
  xlab= "Log2 Fold Change",
  ylab= "Adjusted P-value",
  pCutoff = 0.05,
  FCcutoff= 2.5,
  pointSize= 2.0,
  labSize= 3.0
)

#
EnhancedVolcano(
  seurat.markers,
  lab= rownames(seurat.markers),
  x= "avg_log2FC",
  y= "p_val_adj",
  title = "Enhanced Volcano Plot",
  xlab= "Log2 Fold Change",
  ylab= "Adjusted P-value",
  pCutoff = 10e-16,
  FCcutoff= 0.5,
  pointSize= 3.0,
  labSize= 6.0
)
```


```{r}
saveRDS(seurat_obj, file ="../healthy donors_GSE224995/final_seurat.rds")
```


